Preprocessing and exploration of the narwhal data for the analysis with smoothSDE

### 2024-04-24 R/preprocessing/data_preprocessing.Rmd
changes in tex file and mixed effects analysis
preprocessing updates

llcoord=SpatialPoints(dataprep[,c("Long","Lat")],
                      proj4string=CRS("+proj=longlat +datum=WGS84"))
utmcoord=spTransform(llcoord,CRS("+proj=utm +zone=26 +north,ellps=WGS84"))

@@

  #step lengths
  dx=(sub_data[2:n_sub,"x"]-sub_data[1:(n_sub-1),"x"])/1000
  dy=(sub_data[2:n_sub,"y"]-sub_data[1:(n_sub-1),"y"])/1000

@@

# More data exploration

## Time steps distribution

@@

cat("Percentage of time steps > 2h :",length(delta[delta>2])/length(delta)*100)
```

@@

## Plot all GPS tracks (without interpolation) on a map

@@

```{r histograms distances to shore}
histo1=ggplot(dataprep, aes(x = DistanceShore*1000)) +
  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Original distance to shore (m)")

df=data.frame(DistanceShore=res_coastline$min_distances)
histo2=ggplot(df, aes(x =DistanceShore)) +
  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance from coastline (m)")

df=data.frame(DistanceShore=res_land$min_distances)
histo3=ggplot(df, aes(x =DistanceShore)) +
  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance fom land (m)")
```

@@

cat("Percentage of points on land with distance from land : \n",
    length(which(res_land$min_distances/1000==0))/length(dataprep$time)*100,"\n")

cat("Percentage of points on land with distance from coastline : \n",
    length(which(res_coastline$min_distances/1000==0))/length(dataprep$time)*100)
```

@@

check_plot2=ggplot()+geom_sf(data=land_utm$geometry,fill=NA)+
    coord_sf(datum=st_crs(32626))+
  geom_point(data=df,aes(x,y,col=(DistanceShore==0)),size=0.2)+
    scale_color_viridis_d()

@@

df=data.frame(DistanceShore=res_land$min_distances[res_land$min_distances<100 & res_land$min_distances>0  ])
histo=ggplot(df, aes(x =DistanceShore)) +
  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance to shore (m)")

@@

This seems to be pretty uniformly distributed, so let's choose arbitrarily a threshold of 50 metres.

@@

#replace distance to shore with distances calculated from land polygons (in km)
dataprep$DistanceShore=min_distances/1000

@@

# Convert all UTM coordinates in km

```{r convert to km}

dataprep$px=dataprep$px/1000
dataprep$py=dataprep$py/1000
dataprep$nx=dataprep$nx/1000
dataprep$ny=dataprep$ny/1000

dataprep$x=dataprep$x/1000
dataprep$y=dataprep$y/1000

@@

for (id in unique(dataprep$ID)) {
  sub_data=dataprep[dataprep$ID==id,]
  id_track=ggplot()+geom_sf(data=land_utm$geometry,fill=NA)+
    coord_sf(datum=st_crs(32626))+
  geom_point(data=sub_data,aes(1000*px,1000*py),shape=5,size=0.1,col="red")+
  geom_point(data=sub_data,aes(1000*x,1000*y,col=time),size=0.1)+
    geom_path(data=sub_data,mapping=aes(1000*x,1000*y),linewidth=0.1)+
    geom_point(data = sub_data %>% filter(!duplicated(ID)),
             aes(1000*x,1000*y,col=ID), shape = 3, size = 4,col = "red")+
    scale_color_viridis_c()

@@

ggsave(path=plots_folder,"realExpShore_through_time.png", plot=pExp)
```

@@

```{r plots before and after exposure}
pBE1=milne_land_map + geom_point(dataBE1, mapping=aes(Long, Lat,col=ID),size= 1,alpha=0.8) +
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))

pBE2=milne_land_map + geom_point(dataBE2, mapping=aes(Long, Lat,col=ID),size= 1,alpha=0.8) +
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))

pAE=milne_land_map + geom_point(dataAE, mapping=aes(Long, Lat,col=ID),size= 1,alpha=0.8) +
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))

>>>>

llcoord=SpatialPoints(dataprep[,c("Long","Lat")],
                      proj4string=CRS("+proj=longlat +datum=WGS84"))
utmcoord=spTransform(llcoord,CRS("+proj=utm +zone=26 +north,ellps=WGS84 +units=km"))

@@

  #step lengths
  dx=(sub_data[2:n_sub,"x"]-sub_data[1:(n_sub-1),"x"])
  dy=(sub_data[2:n_sub,"y"]-sub_data[1:(n_sub-1),"y"])

@@

# Time steps distribution

@@

cat("Percentage of time steps > 2h :",length(dtimes$delta[dtimes$delta>2])/length(dtimes$delta)*100)
```

@@

# Time steps distribution during exposure vs no exposure

# Plot all GPS tracks (without interpolation) on a map

@@

```{r histograms distances to shore}
histo1=ggplot(dataprep[dataprep$DistanceShore>0,], aes(x = DistanceShore)) +
  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Original distance to shore (km)")

df=data.frame(DistanceShore=res_coastline$min_distances[res_coastline$min_distances>0])
histo2=ggplot(df, aes(x =DistanceShore)) +
  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance from coastline (km)")

df=data.frame(DistanceShore=res_land$min_distances[res_land$min_distances>0])
histo3=ggplot(df, aes(x =DistanceShore)) +
  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance fom land (km)")
```

@@

cat("Percentage of points on land with distance from land : \n",
    length(which(res_land$min_distances==0))/length(dataprep$time)*100,"\n")

cat("Percentage of points on land with distance from coastline : \n",
    length(which(res_coastline$min_distances==0))/length(dataprep$time)*100)
```

@@

check_plot2=ggplot()+geom_sf(data=land_utm$geometry,fill=NA)+
    coord_sf(datum=st_crs(32626))+
  geom_point(data=df,aes(x,y,col=(DistanceShore<0.1)),size=0.2)+
    scale_color_viridis_d()

@@

df=data.frame(DistanceShore=res_land$min_distances[res_land$min_distances<2 & res_land$min_distances>0  ])
histo=ggplot(df, aes(x =DistanceShore)) +
  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance to shore (km)")

@@

This seems to be pretty uniformly distributed. It is difficult to assess at which distance the narwhals stay from 
the shore in reality.
Let's choose arbitrarily a threshold of 50 metres.

@@

#replace distance to shore with distances calculated from land polygons (in km)
dataprep$DistanceShore=min_distances

@@

@@

for (id in unique(dataprep$ID)) {
  sub_data=dataprep[dataprep$ID==id,]
  id_track=ggplot()+geom_sf(data=land_utm$geometry,fill=NA)+
    coord_sf(datum=st_crs(32626))+
  geom_point(data=sub_data,aes(px,py),shape=5,size=0.1,col="red")+
  geom_point(data=sub_data,aes(x,y,col=time),size=0.1)+
    geom_path(data=sub_data,mapping=aes(x,y),linewidth=0.1)+
    geom_point(data = sub_data %>% filter(!duplicated(ID)),
             aes(x,y,col=ID), shape = 3, size = 4,col = "red")+
    scale_color_viridis_c()

@@

ggsave(path=plots_folder,"realExpShore_through_time.png", plot=pExpShore)
```

@@

```{r plots before and after exposure}
pBE1=milne_land_map + 
  geom_point(dataBE1[,c("ID","Long","Lat","time")],
             mapping=aes(Long, Lat, col = ID),size= 0.5,alpha=1) +
   geom_path(dataBE1[,c("ID","Long","Lat","time")],mapping=aes(Long,Lat,col=ID),linewidth=0.1)+
  geom_point(data = dataBE1[,c("ID","Long","Lat","time")] %>% filter(!duplicated(ID)),
             aes(x = Long, y = Lat,col=ID), shape = 3, size = 4,col = "red")+
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))

pBE2=milne_land_map + 
  geom_point(dataBE2[,c("ID","Long","Lat","time")],
             mapping=aes(Long, Lat, col = ID),size= 0.5,alpha=1) +
   geom_path(dataBE2[,c("ID","Long","Lat","time")],mapping=aes(Long,Lat,col=ID),linewidth=0.1)+
  geom_point(data = dataBE2[,c("ID","Long","Lat","time")] %>% filter(!duplicated(ID)),
             aes(x = Long, y = Lat,col=ID), shape = 3, size = 4,col = "red")+
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))

pAE=milne_land_map + 
  geom_point(dataAE[,c("ID","Long","Lat","time")],
             mapping=aes(Long, Lat, col = ID),size= 0.5,alpha=1) +
   geom_path(dataAE[,c("ID","Long","Lat","time")],mapping=aes(Long,Lat,col=ID),linewidth=0.1)+
  geom_point(data = dataAE[,c("ID","Long","Lat","time")] %>% filter(!duplicated(ID)),
             aes(x = Long, y = Lat,col=ID), shape = 3, size = 4,col = "red")+
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))
